---
title: "SDM Steps"
author: "Alex Musiime"
date: "2025-09-13"
output:
  html_document: default
  html_pdf : default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Objective 1:** Develop high-resolution<br>
dynamic spatial models that describe <br>
the distribution of malaria vector species in Uganda<br>

Plan for objective 1<br>

**Data sets**
1.Collate all vector datasets<br>
2.NMCP datasets<br>
a.year 2020-2026<br>
b.Vector Atlas<br>
c.GBIF<br>
2. Quality check on vector locations<br>
1.Make an intervention dataset space over time<br>
a.?shape file<br>
b.Join the vector dataset?<br>
1.Curate environmental and land cover covariate<br> **raster datasets**
2.MAP standard layers<br>
i.environmental<br>
a.land cover/use (Agriculture)<br>
b.Water Observation from Space (WOfS) long-term frequency on land<br>
-The frequency with each pixel is covered with water<br>
c.Terrain wetness<br>
5 MESS (Multivariate Environmental Similarity Surfaces)<br>
Analysis of Uganda based on a curated set of covariates<br>
1.Consider sampling bias for Uganda<br>
i.accessibility (Weiss - major cities)<br>
ii.burden areas (to highlight in the methods)<br>
iii.intervention areas (to highlight in the methods)<br>
iv.Explore sources of each data sets<br>

**Modelling Approaches**

Write out a mental model of what distribution of malaria<br>
vector species in Uganda (intervention pressure<br> agriculture, environment, sampling bias)<br>
Suggest different models including;<br>
- Generalized Additive Models (GAMs)<br>
-GLM, generalized linear model<br>
-BRT, boosted regression trees<br>
Discuss with supervisors the modelling approaches suggested<br>
Select the best modelling approaches<br>
Fit the model using the different variables<br>
Do cross-validations<br>
Evaluate Metrics<br>
Generate species-specific mean probability or density<br>
Variable importance tables and figures<br>

 **Response**:An_gambiae or An_funestus counts (y)<br> 
 at village level <br>
 or coordinate points<br>
 
varies by each village or coordinate points depending on the covariates<br>
 -climatic variables<br>
 -land cover<br>
 -land use<br>
 -elevation<br>
 -distance to water bodies<br>
 -human population density<br>
 -intervention variables (ITN coverage, IRS coverage)<br>
 -socio-economic variables (wealth index, education level)<br>
 -housing quality (wall type, roof type, floor type)<br>
 -seasonality (wet or dry season)<br>
 -temporal variables (year, month)<br>
 -bias term (travel time to nearest city)<br>
 -incidence/prevalence of malaria<br>
 -intervention variables (ITN coverage, IRS coverage)<br>
 
 Vector Abundance Modelling<br>
  - Poisson GLM <br>
 
$$
Y_i \sim \text{Poisson}(\lambda_i)
$$


$$
\log(\lambda_i) = \beta_{0} + \sum_{j=1}^{k} \beta_{j} X_{ji}
$$
Here<br>
 - \(Y_i\) is the observed count of An_gambiae or An_funestus at location \(i\).<br>

 x, are covariates such as<br>
 household size<br>
 IRS status <br>
 net type <br>
 These models generate predictive abundance maps at<br> household-to-regional scales<br>

Negative binomial regression (for overdispersion):<br>


$$
Y_i \sim \text{NegBin}(\mu_i, \theta)
$$

$$
\log(\mu_i) = \beta_{0} + \sum_{j=1}^{k} \beta_{j} X_{ji}
$$


Model evaluation<br>
 
 -Cross-validation (k-fold, spatial block)<br>
 -Metrics: RMSE, MAE, R² for regression models<br>

Training_data 80%<br>
Test/evaluate 20%<br>


Different Modelling Steps
 
 -Environmental only<br>
 -Environmental +Observation bias<br>
 -Environmental +Observation bias+intervention<br>
 -Environmental +Observation bias+intervention+demography<br>


```{r}
 
```
To account for;<br>
 -Observation bias (travel time to nearest city)<br>
 -Intervention effects (ITN type, IRS status)<br>
 -Demographic factors (household size, net ownership)<br>

Seasonality: Rainfall-driven peaks in vector abundance<br>
IRS decay: Abundance increasing as months since spraying increase<br>
LLIN decay: Declines in net effectiveness with time since<br> distribution campaigns<br>


A generalized additive model (GAM) cAN fit smooth functions <br>
to capture seasonal fluctuations in mosquito counts<br>
 and non-linear effects of environmental covariates<br>
 
 $$
\log(\mu_i) = \beta_{0} + f_{1}(\text{month}_i) + \beta_{1} \,\text{IRS}_i + \beta_{2} \,\text{Nets}_i
$$

where $f_{1}(\text{month}_i)$ captures non-linear seasonal trends.

Hierarchical and Multilevel Models<br>
 
 To account for hierarchical structure of data<br>
 -households nested within villages<br>
 -villages nested within districts<br>
 
 A mixed-effects model with random intercepts for villages<br>
 and districts can capture unobserved heterogeneity at these levels<br>
 

$$
\log(\mu_{ijk}) = \beta_0 + \beta_1 X_{ijk} + u_j + v_{k(j)}
$$

where $u_j$ is a district-level random effect, and $v_{k(j)}$ is a village-level effect nested within district $j$.  

This approach accounts for spatial clustering of mosquito populations, which is critical in malaria ecology.





 



